---
title: "ABS_chi_squared"
author: "EMD"
date: "May 9, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chi-squared tests

STATUS
1. Do the proportions of delta_status for each *taxon* differ from expected?
2. Do the proportions of delta_status for each *realm* differ from expected?

TREND
3. Do the proportions of population trends for each *taxon* differ from expected?
4. Do the proportions of population trends for each *realm* differ from expected?

EXPECTED - informed by IUCN assessments 2007-2018

### What are the expected ratios for each taxa?

#### delta_status

```{r}

library(tidyverse)
#read in file - all assessments 2007-2018 cross-checked for whether they are genuine or not
all_assess <- read.csv(here::here("ChangesInRedList","rl_status_history_2007_2018_genuine.csv"))
str(all_assess)

all_assess_genuine <- all_assess %>% 
  #remove non-genuine changes; keep genuine and NAs (no change)
  filter(is_geniune == "genuine" | is.na(is_geniune)) %>% 
  #add status_impr category column
  mutate(status_impr=case_when( 
    Delta_Status %in% c(-4:-1) ~ "declined",
    Delta_Status == 0 ~ "unchanged",
    Delta_Status %in% c(1:4) ~ "improved")) %>% 
  mutate(status_impr = as.factor(status_impr)) %>% 
  #split mammalia into terrestrial & marine (by family)
  mutate(Taxa = ifelse(Family %in% c("OTARIIDAE","PHOCIDAE","ODOBENIDAE","MUSTELIDAE","URSIDAE","TRICHECHIDAE","DUGONGIDAE","BALAENIDAE","CETOTHERIIDAE","BALAENOPTERIDAE","ESCHRICHTIIDAE","PHYSETERIDAE","KOGIIDAE","MONODONTIDAE","ZIPHIIDAE","DELPHINIDAE","PHOCOENIDAE"),
                       yes = "MAR MAMMALIA",
                       no = Class)) %>% 
  #for some reason, taxa ends up as a number - change back to original factor
  #here, I also combine amphibians and reptiles to be consistent with the apex data groupings
  mutate(Taxa = case_when(Taxa == 2 | Taxa == 6 ~ "REPTAMPH",
                           Taxa == 3 ~ "AVES",
                           Taxa == 5 ~ "MAMMALIA",
                          Taxa == 1 ~ "ACTINOPTERYGII",
                          Taxa == 4 ~ "CHONDRICHTHYES",
                          Taxa=="MAR MAMMALIA" ~ "MAR MAMMALIA"))


#calculate proportions for each taxa
expected_prop <- all_assess_genuine %>% 
  group_by(Taxa) %>% 
  #prop of declined, improved, unchanged and total
  summarize(prop_declined = sum(status_impr== "declined",na.rm=T)/sum(status_impr == "declined" | status_impr == "improved" | status_impr =="unchanged",na.rm=T),
            prop_improved = sum(status_impr== "improved",na.rm=T)/sum(status_impr == "declined" | status_impr == "improved" | status_impr =="unchanged",na.rm=T),
            prop_unchanged = sum(status_impr== "unchanged",na.rm=T)/sum(status_impr == "declined" | status_impr == "improved" | status_impr =="unchanged",na.rm=T)) %>% 
  mutate(Taxa = case_when(Taxa == "AVES" ~ "Birds",
                           Taxa == "ACTINOPTERYGII" ~ "Bony",
                           Taxa == "CHONDRICHTHYES" ~ "Elasmo",
                          Taxa == "MAMMALIA" ~ "LandMam",
                          Taxa == "MAR MAMMALIA" ~"MarineMam",
                          Taxa == "REPTAMPH" ~ "ReptAmph"))
  
expected_prop

#raw numbers & totals, to check math
expected_num <- all_assess_genuine %>% 
  group_by(Taxa) %>% 
  #prop of declined, improved, unchanged and total
  summarize(prop_declined = sum(status_impr== "declined",na.rm=T),
            prop_improved = sum(status_impr== "improved",na.rm=T),
            prop_unchanged = sum(status_impr== "unchanged",na.rm=T),
            total=sum(status_impr == "declined" | status_impr == "improved" | status_impr =="unchanged",na.rm=T))

expected_num

#all spp together - use for realm test
expected_prop_global <- all_assess_genuine %>%
  summarize(prop_declined = sum(status_impr== "declined",na.rm=T)/sum(status_impr == "declined" | status_impr == "improved" | status_impr =="unchanged",na.rm=T),
            prop_improved = sum(status_impr== "improved",na.rm=T)/sum(status_impr == "declined" | status_impr == "improved" | status_impr =="unchanged",na.rm=T),
            prop_unchanged = sum(status_impr== "unchanged",na.rm=T)/sum(status_impr == "declined" | status_impr == "improved" | status_impr =="unchanged",na.rm=T))

expected_prop_global

```

Comprehensively assessed species only
(for the supplement)
https://www.iucnredlist.org/resources/summary-statistics -- see Fig. 2 caption

#### population trend

```{r}
#read in file - rl_summary_2019 (contains popTrend info!)
assess_sum <- read.csv(here::here("data","redlist/rl_summary_2019.csv"))
str(assess_sum)

assess_sum_filtered <- assess_sum %>%
  dplyr::select(phylumName,className,orderName,familyName,genusName,scientificName,redlistCategory,populationTrend) %>% 
  #filter taxa of interest
  filter(className %in% c("MAMMALIA", "AVES","REPTILIA","ACTINOPTERYGII","CHONDRICHTHYES","AMPHIBIA")) %>% 
  #remove NAs & Unknowns for popTrend
  filter (populationTrend %in% c("Decreasing","Increasing","Stable")) %>% 
  droplevels() %>% 
  #split mammalia into terrestrial & marine (by family)
  mutate(Taxa = ifelse(familyName %in% c("OTARIIDAE","PHOCIDAE","ODOBENIDAE","MUSTELIDAE","URSIDAE","TRICHECHIDAE","DUGONGIDAE","BALAENIDAE","CETOTHERIIDAE","BALAENOPTERIDAE","ESCHRICHTIIDAE","PHYSETERIDAE","KOGIIDAE","MONODONTIDAE","ZIPHIIDAE","DELPHINIDAE","PHOCOENIDAE"),
                       yes = "MAR MAMMALIA",
                       no = className)) %>% 
  #for some reason, taxa ends up as a number - change back to original factor
  #here, I also combine amphibians and reptiles to be consistent with the apex data groupings
  mutate(Taxa = case_when(Taxa == 2 | Taxa == 6 ~ "REPTAMPH",
                           Taxa == 3 ~ "AVES",
                           Taxa == 5 ~ "MAMMALIA",
                          Taxa == 1 ~ "ACTINOPTERYGII",
                          Taxa == 4 ~ "CHONDRICHTHYES",
                          Taxa=="MAR MAMMALIA" ~ "MAR MAMMALIA"))


#calculate proportions for each taxa 
expected_prop_popT <- assess_sum_filtered %>% 
  group_by(Taxa) %>% 
  #prop of each popTrend level
  summarize(prop_decrease = sum(populationTrend== "Decreasing",na.rm=T)/sum(populationTrend == "Decreasing" | populationTrend == "Increasing" | populationTrend =="Stable",na.rm=T),
            prop_increase = sum(populationTrend== "Increasing",na.rm=T)/sum(populationTrend == "Decreasing" | populationTrend == "Increasing" | populationTrend =="Stable",na.rm=T),
            prop_stable = sum(populationTrend== "Stable",na.rm=T)/sum(populationTrend == "Decreasing" | populationTrend == "Increasing" | populationTrend =="Stable",na.rm=T)) %>% 
  mutate(Taxa = case_when(Taxa == "AVES" ~ "Birds",
                           Taxa == "ACTINOPTERYGII" ~ "Bony",
                           Taxa == "CHONDRICHTHYES" ~ "Elasmo",
                          Taxa == "MAMMALIA" ~ "LandMam",
                          Taxa == "MAR MAMMALIA" ~"MarineMam",
                          Taxa == "REPTAMPH" ~ "ReptAmph"))

expected_prop_popT

#raw numbers, to check math
expected_num_popT <- assess_sum_filtered %>% 
  group_by(Taxa) %>% 
  #prop of declined, improved, unchanged and total
  summarize(num_decrease = sum(populationTrend== "Decreasing",na.rm=T),
            num_increase = sum(populationTrend== "Increasing",na.rm=T),
            num_stable = sum(populationTrend== "Stable",na.rm=T),
            total=sum(populationTrend == "Decreasing" | populationTrend == "Increasing" | populationTrend =="Stable",na.rm=T))

expected_num_popT

#all spp together - use for realm test
expected_prop_popT_global <- assess_sum_filtered %>%
  summarize(prop_decrease = sum(populationTrend== "Decreasing",na.rm=T)/sum(populationTrend == "Decreasing" | populationTrend == "Increasing" | populationTrend =="Stable",na.rm=T),
            prop_increase = sum(populationTrend== "Increasing",na.rm=T)/sum(populationTrend == "Decreasing" | populationTrend == "Increasing" | populationTrend =="Stable",na.rm=T),
            prop_stable = sum(populationTrend== "Stable",na.rm=T)/sum(populationTrend == "Decreasing" | populationTrend == "Increasing" | populationTrend =="Stable",na.rm=T))

expected_prop_popT_global

```
Comprehensively assessed species only
(for the supplement)
https://www.iucnredlist.org/resources/summary-statistics -- see Fig. 2 caption


### Status

#### By taxon
```{r}
library(tidyverse)
library(here)
library(MASS)

#load in data and explore
#just genuine changes
apex_data_genuine <- read.csv(here::here("data","generated/9_apex_with_gen_changes.csv"))
str(apex_data_genuine)
summary(apex_data_genuine)

```

Re-coded delta_status as improved, unchanged, or declined (status_impr).
```{r}
#re-code delta_status as factor in a new column status_impr (improved, unchanged, declined)
# i.e. -4 through -1 = declined; 0 = unchanged; 1 through 2 = improved

apex_data_genuine <- apex_data_genuine %>% 
  mutate(status_impr=case_when(
    Delta_Status %in% c(-4:-1) ~ "declined",
    Delta_Status == 0 ~ "unchanged",
    Delta_Status %in% c(1:2) ~ "improved")) %>% 
  mutate(status_impr = as.factor(status_impr)) %>% 
  arrange(Delta_Status)

```

Chi-squared test of independence: significant association between status_impr and taxon?
```{r}
tbl_status_taxon = table(apex_data_genuine$status_impr, apex_data_genuine$Taxon)
tbl_status_taxon
chisq.test(tbl_status_taxon) #X-squared = 30.157, df = 10, p-value = 0.0008075
#reject the null that status_impr is independent of taxon. Suggests association (consistent with the lit).

chisq.test(tbl_status_taxon)$expected
#some expected values are <5 so... try Fisher's exact test
fisher.test(tbl_status_taxon,simulate.p.value = T) #p-value = 0.003498

```

Goodness of fit for each taxon: do the distributions differ from expected?
```{r}
levels(apex_data_genuine$status_impr) #"declined"  "improved"  "unchanged" 
taxon_level <- levels(apex_data_genuine$Taxon) #"Birds","Bony","Elasmo","LandMam","MarineMam","ReptAmph"

#filter by birds, as an example
apex_birds <- apex_data_genuine %>% 
  filter(Taxon == "Birds")
#count number of declined, improved, unchanged
bird.d <- sum(apex_birds$status_impr== "declined") #29
bird.i <- sum(apex_birds$status_impr== "improved") #2
bird.u <- sum(apex_birds$status_impr== "unchanged") #79
#test distribution vs. expected (pulled from expected_prop table; taxa names now match apex list)
expected_prop
bird_chi <- chisq.test(c(bird.d,bird.i,bird.u), p = c(expected_prop$prop_declined[expected_prop$Taxa=="Birds"],expected_prop$prop_improved[expected_prop$Taxa=="Birds"],expected_prop$prop_unchanged[expected_prop$Taxa=="Birds"])) #X-squared = 67.534, df = 2, p-value = 2.164e-15
bird_chi
bird_chi$expected
#some expected values <5
chisq.test(c(bird.d,bird.i,bird.u), p = c(expected_prop$prop_declined[expected_prop$Taxa=="Birds"],expected_prop$prop_improved[expected_prop$Taxa=="Birds"],expected_prop$prop_unchanged[expected_prop$Taxa=="Birds"]),simulate.p.value = T)
#X-squared = 67.534, df = NA, p-value = 0.0004998


#for loop saves the day
for(taxa in taxon_level){
  filtered <- subset(apex_data_genuine,Taxon == taxa)
  d <- sum(filtered$status_impr== "declined")
  i <- sum(filtered$status_impr== "improved")
  u <- sum(filtered$status_impr== "unchanged")
  print(taxa)
  print(c("Declined",d)) #observed
  print(c("Improved",i)) #observed
  print(c("Unchanged",u)) #observed
  #pull expected values from expected_prop table
  print(chisq.test(c(d,i,u), p = c(expected_prop$prop_declined[expected_prop$Taxa==taxa],expected_prop$prop_improved[expected_prop$Taxa==taxa],expected_prop$prop_unchanged[expected_prop$Taxa==taxa])))
  print(chisq.test(c(d,i,u), p = c(expected_prop$prop_declined[expected_prop$Taxa==taxa],expected_prop$prop_improved[expected_prop$Taxa==taxa],expected_prop$prop_unchanged[expected_prop$Taxa==taxa]))$expected) #should be >5
  print("Simulated p-values")
  print(chisq.test(c(d,i,u), p = c(expected_prop$prop_declined[expected_prop$Taxa==taxa],expected_prop$prop_improved[expected_prop$Taxa==taxa],expected_prop$prop_unchanged[expected_prop$Taxa==taxa]),simulate.p.value = T)) #if expected <5, use this
}

#Getting error for elasmo (NA p-value due to 0s?) - re-run by hand
apex_elasmo <- apex_data_genuine %>% 
  filter(Taxon == "Elasmo")
#count number of declined, improved, unchanged
elasmo.d <- sum(apex_elasmo$status_impr== "declined") #0
elasmo.i <- sum(apex_elasmo$status_impr== "improved") #1
elasmo.u <- sum(apex_elasmo$status_impr== "unchanged") #46
#test distribution vs. expected (pulled from expected_prop table; taxa names now match apex list)
#change expected for prop_declined and prop_improved from 0 to really low value
elasmo_chi <- chisq.test(c(elasmo.d,elasmo.i,elasmo.u), p = c(0.00000001,0.00000001,0.99999998)) 
elasmo_chi
elasmo_chi$expected #some expected values <5
chisq.test(c(elasmo.d,elasmo.i,elasmo.u), p = c(0.00000001,0.00000001,0.99999998),simulate.p.value = T)
#X-squared = 2127700, df = NA, p-value = 0.0004998

```

#### By realm

Note: the realm_data csv has 555 species in it whereas the 9_apex_with_gen_changes data frame only has 364. Here, I treated the 9_apex_with_gen_changes data frame as the final list. Looks like we're missing data for 34 species.

```{r}
#read in realm data
realm_data <- read.csv(here::here("data","generated/realm_data.csv"))
str(realm_data)

#now I need to add status_impr variable to this data frame
realm_data_short <- realm_data %>% 
  dplyr::select(scientific_name,realm_type,realm) %>% 
  rename(Binomial=scientific_name) 
  
apex_joined <- left_join(apex_data_genuine,realm_data_short) %>%  #longform so there is a row for each spp-realm combo
  mutate(Binomial = as_factor(Binomial))
str(apex_joined) #all spp present (364)
summary(apex_joined) #34 NAs

#omit freshwater realms and drop fw levels (after discussing with Kurt)
#instead, freshwater species have an associated terrestrial realm classification (see realm_data)
apex_joined_no_fw <- apex_joined %>%
 subset(realm_type != "Freshwater") %>% 
  droplevels()

summary(apex_joined_no_fw)
str(apex_joined_no_fw) #249 species, so some are missing
levels(apex_joined_no_fw$realm) #confirm that fw levels have been dropped 

```

Running the chi-squared tests
Note: Since the data frame is longform, some species are counted multiple times if found across more than one realm.

Chi-squared test of independence: significant association between status_impr and realm?
```{r}
tbl_status_realm = table(apex_joined_no_fw$status_impr, apex_joined_no_fw$realm)
tbl_status_realm
chisq.test(tbl_status_realm) 

chisq.test(tbl_status_realm)$expected #some expected values <5, so run Fisher's exact test
fisher.test(tbl_status_realm, simulate.p.value = T) #p-value = 0.004998

```

Goodness of fit for each realm: do the distributions differ from expected?
```{r}
#expected ratios are now drawn from all assessment data

levels(apex_joined_no_fw$status_impr) #"declined"  "improved"  "unchanged" 
realm_level <- levels(apex_joined_no_fw$realm) 
length(realm_level) #16 realms (rather than 29 with fw included)

#for loop
for(r in realm_level){
  filtered <- subset(apex_joined_no_fw,realm == r)
  d <- sum(filtered$status_impr== "declined")
  i <- sum(filtered$status_impr== "improved")
  u <- sum(filtered$status_impr== "unchanged")
  print(r)
  print(c("Declined",d))
  print(c("Improved",i))
  print(c("Unchanged",u))
  print(chisq.test(c(d,i,u), p = c(expected_prop_global$prop_declined,expected_prop_global$prop_improved,expected_prop_global$prop_unchanged))) #equal distribution for geo
  print(chisq.test(c(d,i,u), p = c(expected_prop_global$prop_declined,expected_prop_global$prop_improved,expected_prop_global$prop_unchanged))$expected)
  print("Simulated p-values")
  print(chisq.test(c(d,i,u), p = c(expected_prop_global$prop_declined,expected_prop_global$prop_improved,expected_prop_global$prop_unchanged),simulate.p.value = T)) #if expected <5, use this
}


#check one realm by hand - matches
apex_palearctic <- apex_joined_no_fw %>% 
  filter(realm == "Palearctic")
#count number of declined, improved, unchanged
palearctic.d <- sum(apex_palearctic$status_impr== "declined") #11
palearctic.i <- sum(apex_palearctic$status_impr== "improved") #2
palearctic.u <- sum(apex_palearctic$status_impr== "unchanged") #81
#test distribution vs. expected
chisq.test(c(palearctic.d,palearctic.i,palearctic.u), p = c(expected_prop_global$prop_declined,expected_prop_global$prop_improved,expected_prop_global$prop_unchanged)) 
#X-squared = 172.51, df = 2, p-value < 2.2e-16


```

*A note on test assumptions* 
Expected values for several realms are <5 thus chi-squared approximation may not be correct (one stats website suggested used 2.5 as a cut-off;
https://support.minitab.com/en-us/minitab/18/help-and-how-to/statistics/tables/supporting-topics/chi-square/are-the-results-of-my-chi-square-test-invalid/). Simulating p-values was suggested as a solution. There's also mention of an exact test for goodness of fit, but I'm unable to find suitable r code or recent descriptions of folks using this.

### Population trends

Modified above code for populationTrend variable

```{r}
#remove unknown population trends
apex_data_genuine <- apex_data_genuine %>% 
  filter (populationTrend %in% c("Decreasing","Increasing","Stable")) %>% 
  droplevels()

```


#### By taxon

Chi-squared test of independence: significant association between populationTrend and taxon?
```{r}
tbl_trend_taxon = table(apex_data_genuine$populationTrend, apex_data_genuine$Taxon)
tbl_trend_taxon
chisq.test(tbl_trend_taxon) #X-squared = 61.893, df = 10, p-value = 1.586e-09
#reject the null that populationTrend is independent of taxon. Suggests association (consistent with the lit).
fisher.test(tbl_trend_taxon,simulate.p.value = T) #p-value = 0.0004998

```

Goodness of fit for each taxon: do the distributions differ from expected?
```{r}
levels(apex_data_genuine$populationTrend) #"Decreasing" "Increasing" "Stable" 
taxon_level <- levels(apex_data_genuine$Taxon) #"Birds","Bony","Elasmo","LandMam","MarineMam","ReptAmph" 
expected_prop_popT #expected

#for loop
for(taxa in taxon_level){
  filtered <- subset(apex_data_genuine,Taxon == taxa)
  d <- sum(filtered$populationTrend== "Decreasing")
  i <- sum(filtered$populationTrend== "Increasing")
  s <- sum(filtered$populationTrend== "Stable")
  print(taxa)
  print(c("Decreasing",d))
  print(c("Increasing",i))
  print(c("Stable",s))
  #pull from expected_prop_popT table
  print(chisq.test(c(d,i,s), p = c(expected_prop_popT$prop_decrease[expected_prop_popT$Taxa==taxa],expected_prop_popT$prop_increase[expected_prop_popT$Taxa==taxa],expected_prop_popT$prop_stable[expected_prop_popT$Taxa==taxa])))
  print(chisq.test(c(d,i,s), p = c(expected_prop_popT$prop_decrease[expected_prop_popT$Taxa==taxa],expected_prop_popT$prop_increase[expected_prop_popT$Taxa==taxa],expected_prop_popT$prop_stable[expected_prop_popT$Taxa==taxa]))$expected)
  print("Simulated p-values")
  print(chisq.test(c(d,i,s), p = c(expected_prop_popT$prop_decrease[expected_prop_popT$Taxa==taxa],expected_prop_popT$prop_increase[expected_prop_popT$Taxa==taxa],expected_prop_popT$prop_stable[expected_prop_popT$Taxa==taxa]),simulate.p.value = T))
}

#hand check one (birds)
apex_birds <- apex_data_genuine %>% 
  filter(Taxon == "Birds")
#count number of declined, improved, unchanged
bird.d.trend <- sum(apex_birds$populationTrend== "Decreasing") #65
bird.i.trend <- sum(apex_birds$populationTrend== "Increasing") #14
bird.s.trend <- sum(apex_birds$populationTrend== "Stable") #30
#test distribution vs. expected
chisq.test(c(bird.d.trend,bird.i.trend,bird.s.trend), p = c(expected_prop_popT$prop_decrease[expected_prop_popT$Taxa=="Birds"],expected_prop_popT$prop_increase[expected_prop_popT$Taxa=="Birds"],expected_prop_popT$prop_stable[expected_prop_popT$Taxa=="Birds"])) 
#X-squared = 2.5757, df = 2, p-value = 0.2759


```

#### By realm

Chi-squared test of independence: significant association between populationTrend and realm?
```{r}
apex_joined_no_fw <- apex_joined_no_fw %>% 
  filter (populationTrend %in% c("Decreasing","Increasing","Stable")) %>% 
  droplevels()

tbl_trend_realm = table(apex_joined_no_fw$populationTrend, apex_joined_no_fw$realm)
tbl_trend_realm
chisq.test(tbl_trend_realm) #X-squared = 86.688, df = 30, p-value = 2.068e-07
#reject null hypothesis that populationTrend is independent of realm (same conclusion as with fw)

fisher.test(tbl_trend_realm,simulate.p.value = T) #p-value = 0.0004998
```

Goodness of fit for each realm: do the distributions differ from expected?
```{r}
#expected ratios are now drawn from all assessment data

levels(apex_joined_no_fw$populationTrend) 
realm_level <- levels(apex_joined_no_fw$realm) 

#for loop
for(r in realm_level){
  filtered <- subset(apex_joined_no_fw,realm == r)
  d <- sum(filtered$populationTrend== "Decreasing")
  i <- sum(filtered$populationTrend== "Increasing")
  s <- sum(filtered$populationTrend== "Stable")
  print(r)
  print(c("Decreasing",d))
  print(c("Increasing",i))
  print(c("Stable",s))
  print(chisq.test(c(d,i,s), p = c(expected_prop_popT_global$prop_decrease,expected_prop_popT_global$prop_increase,expected_prop_popT_global$prop_stable))) #equal distribution for geo
  print(chisq.test(c(d,i,s), p = c(expected_prop_popT_global$prop_decrease,expected_prop_popT_global$prop_increase,expected_prop_popT_global$prop_stable))$expected) #expected values; should be above 5
  print("Simulated p-values")
  print(chisq.test(c(d,i,s), p = c(expected_prop_popT_global$prop_decrease,expected_prop_popT_global$prop_increase,expected_prop_popT_global$prop_stable),simulate.p.value = T))
}

#check one realm by hand - matches
apex_palearctic <- apex_joined_no_fw %>% 
  filter(realm == "Palearctic")
#count number of declined, improved, unchanged
palearctic.d <- sum(apex_palearctic$populationTrend== "Decreasing") 
palearctic.i <- sum(apex_palearctic$populationTrend== "Increasing") 
palearctic.s <- sum(apex_palearctic$populationTrend== "Stable") 
#test distribution vs. expected
chi <- chisq.test(c(palearctic.d,palearctic.i,palearctic.s), p = c(expected_prop_popT_global$prop_decrease,expected_prop_popT_global$prop_increase,expected_prop_popT_global$prop_stable)) 
chi #X-squared = 34.978, df = 2, p-value = 2.539e-08
chi$expected 
chi$observed 

```
